Modelling Predator Preferences using Expectation-Maximization Algorithm and Molecular Gut-Content Analysis

### 2015-01-12 paper/data.tex
first bio comments taken. need more references...

\section{Real Data}
\label{sec:data}

We analyzed a dataset that was collected to investigate the eating preferences of the wolf spider, genus \textit{Schizocosa}, towards the prey orders Diptera and Collembola.  Predators were collected in traps within the deciduous Berea College Forest in Madison County, Kentucky, USA and had their gut-contents analyzed to determine whether or not the spiders ate any Diptera or Collembola within each time period.  The left panel of figure~\ref{fig:data} plots the percentage of spiders that had the prey in their guts across time.  Prey were also collected in traps from the same area.  These data were collected for each month from October $2011$ to March $2013$.  On average, $69$ spiders, and $111$ and $297$ Diptera and Collembola, respectively, were caught in each time period.  The range of the sample sizes, across all $18$ months is, $11$ to $181$ for caught spiders, from $7$ to $322$ for trapped Diptera, and from $101$ to $755$ for trapped Collembola.  The right panel of figure~\ref{fig:data} plots the total number of each order that was caught during each time period.

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{data}
  \caption{For both Diptera and Collembola, the plots show the proportion of spiders with these orders found in their guts, and the number of these orders trapped in each time period.}
  \label{fig:data}
\end{figure}

These data provide an example of our hierarchy of hypotheses.  First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders Diptera  and Collembola.  With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.  Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.  Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.  In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, $p-value < 0.0001$.  Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.  Figure~\ref{fig:cst} plots the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{cst}
  \caption{One plot for each prey order displays the point estimates and $95\%$ confidence intervals as estimated from the model $c_{st}$.}
  \label{fig:cst}
\end{figure}

>>>>

\section{Application}
\label{sec:data}

We analyzed a dataset that was collected to investigate the feeding preferences of two species of wolf spider, Schizocosa ocreata and Schizocosa stridulans (Araneae: Lycosidae).  Every $6-12$ days, $10$ to $40$ spiders were hand-collected between October $2011$ and April $2013$ within Berea College Forest in Madison County, Kentucky, USA.  Spiders were removed from the litter using an aspirator, placed in separate $1.5$ mL microcentrifuge tubes filled with $95$\% {\color{red}EtOH}, and preserved at $-20\celsius$ until DNA extraction.  In parallel, we also surveyed availability of forest floor prey using pitfall traps $(n = 32)$.  For the analysis, both species of Schizocosa were pooled and the number of spiders and prey were analysed by month.  On average, $69$ spiders, $111$ Diptera, and $297$ Collembola were caught in each time period.  The range of the sample sizes across all $18$ months was $11$ to $181$ for caught spiders, $7$ to $322$ for trapped Diptera, and $101$ to $755$ for trapped Collembola.  Figure~\ref{fig:trapped} plots the total number of each order that was caught during each time period.

\begin{table}
  \label{tab:s1}
  \centering
  \begin{tabular}{llll}
    \hline
    \textbf{Target group} & \textbf{Primer names and sequences} $5'-3'$ & \textbf{Size (bp)} & \textbf{Source}\\
    \hline
    Collembola & Col3F: GGACGATYTTRTTRGTTCGT & 228 & Sint et al. 2012 \\
    & Col-gen-A246: TTTCACCTCTAACGTCGCAG & & \\
    Diptera & DIPS16: CACTTGCTTCTTAAATrGACAAATT & 198 & Eitzinger et al. 2013 \\
    & DIPA17: TTyATGTGAACAGTTTCAGTyCA &  & \\
  \end{tabular}
  \caption{Targeted prey orders, primer names and sequences, size of amplicon, and source of design for the detection of prey taxa within the guts of Schizocosa spiders.  Both primer sets were used in singleplex PCR assays.}
\end{table}

To determine whether spiders consumed dipterans and/or collemblans, we conducted a molecular analysis of their gut-contents.  First, DNA from spiders was extracted using Qiagen DNEasy\circledR Tissue Extraction Kit {\color{red}(Qiagen Inc., Chatsworth, California, USA)} following the animal tissue protocol outlined by the manufacturer, with minor modifications.  Whole bodies of the spiders were first crushed to release prey DNA from within their alimentary canal for extraction.  The $200 \mu$L extractions were stored at $-20\celsius$ until PCR.  Second, order-specific primers from the literature were used to detect the DNA of Collembola and Diptera within the guts of the spiders.  Primer pairs designed by {\color{red} Sint et al. (2012)}, targeting the $18$S rDNA gene, were used to detect Collembola predation table~\ref{tab:s1}.  A PCR cycling protocol for $12.5 \mu$L reactions containing $1x$ Takara buffer {\color{red}(Takara Bio Inc., Shiga, Japan)}, $0.2$ mM dNTPs, $0.2 \mu$M of each primer, $0.625$ U Takara Ex TaqTM and $1.5 \mu$L of template DNA, using BioRad PTC$-200$ and C$1000$ thermal cyclers {\color{red}(Bio-Rad Laboratories, Hercules, California, USA)}, was optimized as follows: $95\celsius$ for $1$ minute, followed by $35$ cycles of $94\celsius$ for $30$ seconds, $61.2\celsius$ for $90$ seconds, and $72\celsius$ for $60$ seconds.  Primer pairs designed by {\color{red}Eitzinger et al. (2013)}, targeting the $18$S rDNA gene, were used to detect Diptera predation table~\ref{tab:s1}.  PCR cycling protocol for $12.5 \mu$L reactions with Takara reagents (as above) and $2 \mu$L of template DNA was optimized as follows: $95\celsius$ for $1$ minute, followed by $40$ cycles of $94$ for $45$ seconds, $60\celsius$ for $45$ seconds, and $72\celsius$ for $45$ seconds.  Both primer pairs were tested for cross-reactivity against a range of prey and predator species from the field site and in all cases, no amplification of DNA was observed, confirming suitable specificity of the primers for this study.  Lastly, electrophoresis of $10 \mu$L of each PCR product was later conducted to determine success of DNA amplification using $2\%$ Seakem agarose {\color{red}(Lonza, Rockland, Maine, USA)} stained with $1x$ GelRed\texttrademark nucleic acid stain {\color{red}(Biotium, Hayward, California, USA)}.  This procedure allowed us to determine a presence or an absence of Diptera and Collembola DNA within each spider.

\begin{figure}
  \centering
  \includegraphics[scale=0.35]{prey_trapped}
  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the number of the prey trapped in each time period.}
  \label{fig:trapped}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.35]{prey_props}
  \caption{For both Collembola (pink/dashed) and Diptera (blue/solid), the plot shows the prey proportions in the sampled wolf spiders' guts in each time period.}
  \label{fig:props}
\end{figure}

These data provide an example of our hierarchy of hypotheses.  First, we tested model $c_{st} = c$ against $c_{st} = c_s$, to determine whether or not the wolf spider has different preferences for the two orders Diptera  and Collembola.  With, one degree of freedom, this likelihood ratio test indicated, $p-value < 0.0001$,  that two parameters, one for each order, fits these data better than one parameter for both.  Similarly, we tested whether or not there was a significant effect across time by testing model $c_{st} = c$ against $c_{st} = c_t$.  Here, the likelihood ratio test, with $17$ degrees of freedom, implies that the wolf spiders of the Berea College Forest eat these prey orders at different rates across the months of the year, $p-value < 0.0001$.  In fact, we find that the most parameter rich model, $\lambda_{st} = c_{st} \gamma_{st}$ fits these data better than is expected by chance, $p-value < 0.0001$.  Model $c_{st}$ estimates $72$ parameters in total; since, in this case, there are two prey of interest and $18$ time periods, it takes $36$ parameters to estimate each $c_{st}$ and $\gamma_{st}$.  Figures~\ref{fig:est_coll},~\ref{fig:est_dipt} plot the point estimates and $95\%$ confidence intervals of $c_{st}$, for both prey across all time periods.

\begin{figure}
  \centering
  \includegraphics[scale=0.35]{est_coll}
  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Collembola.}
  \label{fig:est_coll}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.35]{est_dipt}
  \caption{Point estimates (blue/solid) and $95\%$ confidence intervals (pink/dashed) as estimated from the model $c_{st}$ for Diptera.}
  \label{fig:est_dipt}
\end{figure}

### 2015-01-12 paper/discussion.tex
first bio comments taken. need more references...

The model developed here allows predators' preferences by testing simultaneously across an array of multiple prey species and time points.  This is achieved via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the relative preferences for different species.

@@

Further developments of our model could be beneficial.  Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature might be advantageous.

>>>>

The model developed here allows for the determination of a predator's preferences by testing simultaneously across an array of multiple prey species and time points.  This is achieved via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, or relative rates across species groups allows for statements about the relative preferences for different species.

@@

Further developments of our model could be beneficial.  Taking into account other environmental variables that might effect a predator's eating habits, such as rain or temperature might also be advantageous.

### 2015-01-12 paper/introduction.tex
first bio comments taken. need more references...

The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them; a quick search of just $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigor of a model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  We propose an intuitive statistical model to determine and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

A comprehensive overview by \citet{Lechowicz:1982}, which was later summarized by \citet{Manly:1992}, details the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.  While \citet{Lechowicz:1982} recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of suggesting the indices which  estimate biologically meaningful values.  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

To encourage more formal statistical inference, and simultaneously generalize predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalized linear models (GLM).  The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species to that which is available while using environmental variables as predictors.  The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.  Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.  To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.  An outcome of our model, is that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.

Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.  This provides researchers a more detailed analysis of the predator's eating preferences.  Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact tallies of the number of each prey species eaten within any given time period is not observed.  Instead, we rely on the researcher being able to DNA sequence the contents of the predator's gut and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

This paper is organized as follows.  Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the Expectation-Maximization (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.  In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.  Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of the wolf spider, found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how interested practitioners could apply our methods.  A brief discussion concludes the paper in section~\ref{sec:discussion}.  Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that fits all the methods discussed.

>>>>

The indices most commonly used to estimate a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them; a quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  We propose an intuitive statistical model to determine and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

A comprehensive overview by \citet{Lechowicz:1982}, which was later summarized by \citet{Manly:1992}, details the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save Strauss' linear index $L$, despite the fact that most of the methods differ by range and linearity of response.  While \citet{Lechowicz:1982} recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best,'' albeit imperfect, index, \citet{Manly:1992} instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value.''  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and because the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species to that which is available while using environmental variables as predictors.  The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.  Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.  To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.  An outcome of our model, is that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.

Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.  This provides researchers a more detailed analysis of the predator's feeding preferences.  Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact tallies of the number of each prey species eaten within any given time period is not observed.  Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

This paper is organized as follows.  Section~\ref{sec:methods} describes our statistical model, for both fully observed count data, and for the non-observed count data for which we use the Expectation-Maximization (EM) algorithm, and the statistical tests used to make statements about the population parameters of interest.  In section~\ref{sec:sim}, we offer a simulation study that demonstrates the accuracy of our methods.  Section~\ref{sec:data} provides a real dataset, which investigates the eating preferences of wolf spiders (Araneae: Lycosidae), found in the Berea College Forest in Madison County, Kentucky, USA, to demonstrate how those interested in assessing trophic interactions with gut-content analyses could apply our methods.  A brief discussion concludes the paper in section~\ref{sec:discussion}.  Alongside our model, we offer an \texttt{R} \citep{Core-Team:2014} package named \texttt{spiders} that fits all the methods discussed.

### 2015-01-12 paper/main.tex
first bio comments taken. need more references...

\begin{document}

\title{Modeling Predator Preferences}
\author{Edward A. Roualdes and Simon Bonner \\ University of Kentucky}
\date{\today}
\maketitle

\begin{abstract}
The literature on modeling a predator's prey selection includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.  Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.  Further, we apply the Expectation-Maximization algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.  A simulation is provided to demonstrate the accuracy of our methods.
\end{abstract}

\input{introduction.tex}
\input{methods.tex}
\input{simulations.tex}
\input{data.tex}
\input{discussion.tex}

>>>>

\begin{document}
\subsection*{Title:}  Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

\subsection*{Author Information:}
Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^3$, and James D. Harwood$^4$

\begin{tabular}{ll}
  1 & Department of Statistics\\
  & University of Kentucky\\
  & Rm.~311 Multidisciplinary Science Building\\
  & 725 Rose Street\\
  & Lexington, KY 40536-0082\\
  & USA\\
  \\
  2 & Email: edward.roualdes@uky.edu\\
  & Phone: 530-570-2674\\
  \\
  3 & Warnell School of Forestry and Natural Resources \\
  & University of Georgia \\
  & Athens, GA 30602\\
  & USA \\
  \\
  4 & Department of Entomology\\
  & University of Kentucky \\
  & Lexington, KY 40546\\
  & USA \\
\end{tabular}

\subsection*{Short Title:}
Modelling Predator Preferances

\subsection*{Word Count:}
This manuscript contains $3253$ words including figure captions and headers, but excluding title page and equations.

\subsection*{Keywords}
electivity; Expectation-Maximization;

\newpage
\section*{Summary}
\label{sec:abstract}

\begin{enumerate}
\item The literature on modelling the prey selection a predator includes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of multiple species and time points.
\item Further, we apply the Expectation-Maximisation algorithm to cases where exact counts of the number of prey species eaten in a particular timer period is not observed.
\item We conduct a simulation study to demonstrate the accuracy of our method, and a real dataset, collected on wolf spiders using molecular gut-content analysis, illustrates how to apply our methods.
\end{enumerate}

\input{introduction.tex}
\input{methods.tex}
\input{simulations.tex}
\input{data.tex}
\input{discussion.tex}
\input{acknowledgments.tex}
\input{accessibility.tex}

### 2015-01-12 paper/methods.tex
first bio comments taken. need more references...

The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat prey species independently, $3)$ predators eat independent of each other.

@@

The first hypothesis states that predators and traps sample all prey species at the same rate.  One imagines this is the case if the predator simply eats that which comes within its reach, thus suggesting a diet indifference.  The second says that predators sample prey proportionally across all time periods.  The third hypothesis says that predators sample different prey species at different rates, but each rate is steady across time.  This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.  Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.  The fifth assumes a predator's selection varies by both time and prey species.  This would make sense if environmental variables, say weather, or prey availability, and taste were affecting predators' selection strategies.

@@

  \caption{Hierarchy of hypotheses.}
  \label{fig:hier}
\end{figure}

@@

The likelihood function that allows for estimation of these parameters is as follows.  Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to get the likelihood.

@@

Many times it may not be possible to count the number of individuals of each prey species that are in a predator's gut.  Instead, it may only be possible to detect whether or not a predator consumed the prey species during a given period.  In this case we can still make inference about the predators' preferences for the different prey species by using the Expectation-Maximization (EM) algorithm to compute maximum likelihood estimates.

@@

The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
\begin{align}
  \label{eq:estep}
  \begin{split}
  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
    + z_{jst} \log{\lambda_{st}} \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log{\lambda_{st}} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
\end{split}
\end{align}
\noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.  In fact, as we only need find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.  Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.

As we show in our simulation study, this generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.  In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.  On the other hand, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.  The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} 1 - \exp\{-\lambda_{st}\}$.  As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.  When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.

\subsection{Testing}

The likelihood ratio test statistic is

>>>>

The use of Poisson distributions make the following implicit assumptions: $1)$ traps independently catch the prey species of interest, $2)$ predators eat independent of each other.

@@

The first hypothesis states that predators and traps sample all prey species at the same rate.  One imagines this is the case if the predator simply eats prey which comes within its reach, thus suggesting no selection for a particular prey item.  The second states that predators sample prey proportionally across all time periods.  The third hypothesis states that predators sample different prey species at different rates, but each rate is steady across time.  This implies that the predator expresses preferences for one prey species over another, but is unresponsive to changes due to time.  Conversely, the forth hypothesis implies that each prey species is sampled similarly within each time period, while the rates across time are allowed to change.  The fifth hypothesis assumes a predator's selection varies by both time and prey species.  This would make sense if environmental and biological variables, such as weather, prey availability, and/or palatability were affecting predators' selection strategies.

@@

  \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.  One begins with the models at the top and sequentially, following the arrows, tests the hypotheses using the formal test described in section~\ref{sec:test}.}
  \label{fig:hier}
\end{figure}

@@

The likelihood function that allows for estimation of these parameters is as follows.  Since we assume $X_{jst}$ is independent of $Y_{ist}$ we can simply multiply the respective Poisson probability density functions, and then form products over all $s,t$ to obtain the likelihood.

@@

In many applications, such as DNA-based gut-content anslysis, it is not possible to count the number of individuals of each prey species that are in a predator's gut.  Instead, it is only possible to detect whether or not a predator consumed the prey species during a given time period, based on the rate at which prey DNA decays in the predator gut \citep{Greenstone:2013}.  In this case we can still make inference about the predators' preferences for the different prey species by using the Expectation-Maximization (EM) algorithm to compute maximum likelihood estimates.

@@

The calculation of $Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})$ is not difficult and is given by:
\begin{align}
  \label{eq:estep}
  \begin{split}
  Q^{(k)}(\boldsymbol{\lambda}, \boldsymbol{\gamma})
  & = \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})} + \log{f_{Y|\boldsymbol{\gamma}}(y_{ist})} \\
  & = \sum_{s=1}^S \sum_{t=1}^T \sum_{j=1}^{J_t} \E \log{f_{X,Z|\boldsymbol{\lambda}}(X_{jst},z_{jst})}
  + \sum_{s=1}^S \sum_{t=1}^T \sum_{i=1}^{I_t} \log{f_{Y|\boldsymbol{\gamma}}(y)} \\
  & \propto \sum_{s,t,j} \left( - \lambda_{st} 
    + z_{jst} \log{\lambda_{st}} \E X_{jst} \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right) \\
  & \propto \sum_{s,t} \left( -J_t \lambda_{st} + z_{\cdot st} \log{\lambda_{st}} \E (X_{jst}|\lambda_{st}^{(k)},\gamma_{st}^{(k)}) \right) + \sum_{s,t} \left( -I_t \gamma_{st} + Y_{\cdot st} \log{I_t \gamma_{st}} \right).
\end{split}
\end{align}
\noindent No analytic solution to the M-step exists, however, so we chose to maximize $Q$ with coordinate descent \citep{Luo:1992}.  In fact, as we only need to find parameters that increase the value of $Q$ on each iteration, we forgo fully iterating to find the maximum and instead perform just one step uphill within each EM iteration.  Since $Q^{(k)}$ is concave and smooth in the parameters $\boldsymbol{\lambda}, \boldsymbol{\gamma}$, we are able to use the convergence of parameter estimates, $||(\boldsymbol{\lambda}^{(k)}, \boldsymbol{\gamma}^{(k)}) - (\boldsymbol{\lambda}^{(k+1)}, \boldsymbol{\gamma}^{(k+1)})||_{\infty} < \tau $, for some $\tau>0$, as our stopping criterion.

As we show in our simulation study, this generalized EM algorithm accurately estimates the parameters when values of $\lambda_{st}$ are relatively small, such that zeros are prevalent in the data $Z_{jst}$.  In this case, not too much information is lost since estimation of $\E Z_{jst}$ can be estimated well by the proportion of observed zeros.  In contrast, if the predator consistently eats a given prey species, few to no zeros will show up in the observed data and $\E Z_{jst}$ is estimated to be nearly $1$.  The loss of information is best seen by attempting to solve for $\lambda_{st}$ in the equation $1 = \E Z_{jst} = 1 - \exp\{-\lambda_{st}\}$.  As the proportion of ones in the observed data increases, we expect $\lambda_{st}$ to grow exponentially large.  When no zeros are present in the data, so that where only ones are observed, the likelihood can be made arbitrarily large by sending the parameter off to infinity.

\subsection{Testing}
\label{sec:test}
The likelihood ratio test statistic is
